---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CausalMultiOmics

<!-- badges: start -->
<!-- badges: end -->

CausalMultiOmics audits multi-block datasets before they are integrated.

Data blocks of any modality — omics, imaging, clinical, environmental,
wearable devices — are collected into a single container and profiled block
by block: structural problems, missing-value patterns, distributional shape,
feature quality and multivariate structure. For every block a battery of
candidate transformations is benchmarked and an automatic preprocessing
recipe is derived from the result, together with a quality score,
quality-control checks, diagnostic plots and a self-contained HTML report.

Nothing is modified in place. `check_data()` never touches its input; every
finding is returned inside a single object.

## Installation

CausalMultiOmics is not on CRAN yet. Install it from a local checkout:

``` r
# install.packages("devtools")
devtools::install("path/to/CausalMultiOmics")
```

## Example

Blocks are supplied as a named list. They do not need to share samples, but
each one must carry its sample identifiers in its row names — that is what
links blocks to each other and to the metadata.

```{r example}
library(CausalMultiOmics)

set.seed(1)

transcriptomics <- data.frame(
  G1 = rnorm(20),
  G2 = rnorm(20),
  G3 = c(rnorm(19), NA),
  row.names = paste0("S", 1:20)
)

proteomics <- data.frame(
  P1 = rpois(18, 5),
  P2 = rpois(18, 10),
  row.names = paste0("S", 3:20)
)

omics <- load_data(
  assays = list(
    transcriptomics = transcriptomics,
    proteomics = proteomics
  ),
  metadata = data.frame(
    sample_id = paste0("S", 1:20),
    group = rep(c("A", "B"), each = 10),
    stringsAsFactors = FALSE
  )
)

omics
```

`check_data()` performs the audit and returns a `CMOValidation` object.

```{r validate}
validation <- check_data(omics)

validation
```

Each block gets its own diagnostics, transformation benchmark and
preprocessing recipe:

```{r recipe}
validation$diagnostics$proteomics$data_type

validation$transformations$proteomics$recommended

validation$recipes$proteomics$imputation
```

`preprocess()` executes that plan and records every step, so the pipeline can
be replayed on a second cohort with `apply_preprocessing()`.

```{r preprocess}
clean <- preprocess(omics, validation, plots = FALSE, quiet = TRUE)

clean
```

## Building evidence

`analyze()` does not fit a model and hand it back. It runs every applicable
method, collects the relationships each one reports, and merges them into a
single scored graph.

```{r analyze}
results <- analyze(clean, outcome = "group", effort = "fast",
                   plots = FALSE, quiet = TRUE)

results
```

Every relationship carries the identification strategy that would license
reading it causally, and the assumptions that would have to hold. An edge
identified as `"adjustment"` is an association however high it scores.

```{r evidence}
results$graph$edges[, c("source", "target", "level_label",
                        "identification", "evidence_score")]
```

`explain()` gathers everything known about one variable, including what each
method reported on its own before anything was merged.

```{r explain, eval = FALSE}
explain(results, "P1")
```

`counterfactual()` states the findings in the units the variable was measured
in. Nothing is included unless you name it as something that could plausibly
be changed &mdash; a contrast for a genotype is arithmetic without meaning.

```{r counterfactual, eval = FALSE}
counterfactual(results, modifiable = "proteomics")
```

Prior biological knowledge can be attached with `annotate_evidence()`. It is
reported beside the score and never folded into it, so a relationship nobody
has published yet is not penalised for being new. It takes a local export
rather than a live query, which keeps the analysis offline, deterministic and
independent of a database version the result could not record.

```{r annotate, eval = FALSE}
known <- data.frame(source = "P1", target = "M1",
                    support = 0.9, database = "local STRING export")

annotate_evidence(results, known)
```

## Reading the results

`report()` turns a validation into something readable. The console format
leads with the verdict and flags only what needs attention:

```{r report, eval = FALSE}
report(validation, format = "console")
```

The default format writes a self-contained HTML document — styles, scripts
and plots are all inlined, so it opens with no network access and no CDN
dependency:

```{r report-html, eval = FALSE}
report(validation, file = "validation_report.html")
```

## Development

```{r dev, eval = FALSE}
devtools::document()
devtools::test()
devtools::check()
```

`tests/manual/test_all.R` is a standalone development harness that exercises
the internals and prints what a user would see at the console. It is not part
of the built package.
