---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CausalMultiOmics

<!-- badges: start -->
<!-- badges: end -->

CausalMultiOmics audits multi-block datasets before they are integrated.

Data blocks of any modality — omics, imaging, clinical, environmental,
wearable devices — are collected into a single container and profiled block
by block: structural problems, missing-value patterns, distributional shape,
feature quality and multivariate structure. For every block a battery of
candidate transformations is benchmarked and an automatic preprocessing
recipe is derived from the result, together with a quality score,
quality-control checks, diagnostic plots and a self-contained HTML report.

Nothing is modified in place. `check_data()` never touches its input; every
finding is returned inside a single object.

## Installation

CausalMultiOmics is not on CRAN yet. Install it from a local checkout:

``` r
# install.packages("devtools")
devtools::install("path/to/CausalMultiOmics")
```

## Example

Blocks are supplied as a named list. They do not need to share samples, but
each one must carry its sample identifiers in its row names — that is what
links blocks to each other and to the metadata.

```{r example}
library(CausalMultiOmics)

set.seed(1)

transcriptomics <- data.frame(
  G1 = rnorm(20),
  G2 = rnorm(20),
  G3 = c(rnorm(19), NA),
  row.names = paste0("S", 1:20)
)

proteomics <- data.frame(
  P1 = rpois(18, 5),
  P2 = rpois(18, 10),
  row.names = paste0("S", 3:20)
)

omics <- load_data(
  assays = list(
    transcriptomics = transcriptomics,
    proteomics = proteomics
  ),
  metadata = data.frame(
    sample_id = paste0("S", 1:20),
    group = rep(c("A", "B"), each = 10),
    stringsAsFactors = FALSE
  )
)

omics
```

`check_data()` performs the audit and returns a `CMOValidation` object.

```{r validate}
validation <- check_data(omics)

validation
```

Each block gets its own diagnostics, transformation benchmark and
preprocessing recipe:

```{r recipe}
validation$diagnostics$proteomics$data_type

validation$transformations$proteomics$recommended

validation$recipes$proteomics$imputation
```

`preprocess()` executes that plan and records every step, so the pipeline can
be replayed on a second cohort with `apply_preprocessing()`.

```{r preprocess}
clean <- preprocess(omics, validation, plots = FALSE, quiet = TRUE)

clean
```

## Building evidence

`analyze()` does not fit a model and hand it back. It runs every applicable
method, collects the relationships each one reports, and merges them into a
single scored graph.

```{r analyze}
results <- analyze(clean, outcome = "group", effort = "fast",
                   plots = FALSE, quiet = TRUE)

results
```

Every relationship carries the identification strategy that would license
reading it causally, and the assumptions that would have to hold. An edge
identified as `"adjustment"` is an association however high it scores.

```{r evidence}
results$graph$edges[, c("source", "target", "level_label",
                        "identification", "evidence_score")]
```

`explain()` gathers everything known about one variable, including what each
method reported on its own before anything was merged.

```{r explain, eval = FALSE}
explain(results, "P1")
```

## What stops a number meaning more than it should

Four checks run over the merged graph. None of them changes an estimate;
they change what may be claimed about one, and what it is ranked by.

**The quantity each method measured.** Only quantities of the same family are
pooled. Averaging a log hazard ratio with a correlation coefficient produced
a headline number in no units at all, and it was the number the report led
with.

**A causal structure, if you have one.** A DAG is a claim about how the world
works, not something recoverable from a correlation matrix, so it has to come
from you. Supplying one turns "we adjusted for age" into a verdict:

```{r dag, eval = FALSE}
structure <- data.frame(
  from = c("age", "age", "protein", "inflammation"),
  to   = c("protein", "disease", "inflammation", "disease")
)

# Adjusting for the mediator removes part of the effect being measured.
check_dag(structure, "protein", "disease", adjusted = c("age", "inflammation"))

analyze(clean, outcome = "group", dag = structure)
```

Adjusting for a mediator or a collider makes an estimate worse than doing
nothing. Both look like diligence, and only a declared structure tells them
apart.

**How much of it was measured.** `data_quality` is the share of the data
behind a relationship that was measured rather than reconstructed. Imputation
fills gaps with plausible numbers, and from that moment nothing downstream
can tell a measurement from an estimate: a variable that arrived a quarter
empty reports the same precision as one fully observed. Only imputed values
count against it, since rows dropped for being incomplete cost sample size
and that is already in the precision score.

**Whether the picture repeats.** `result$consensus` is the graph across
resamples rather than in the one sample collected. Per-edge stability answers
"would this relationship come back?" one at a time; it cannot answer "would
this picture come back?". It also says what it is not: a bootstrap resamples
the people who were measured, so it reports how much the result depends on
which of them ended up in the study, never whether a relationship is real.

```{r consensus, eval = FALSE}
results <- analyze(clean, outcome = "group", effort = "standard")

results$consensus
```

`counterfactual()` states the findings in the units the variable was measured
in. Nothing is included unless you name it as something that could plausibly
be changed &mdash; a contrast for a genotype is arithmetic without meaning.

```{r counterfactual, eval = FALSE}
counterfactual(results, modifiable = "proteomics")
```

Prior biological knowledge can be attached with `annotate_evidence()`. It is
reported beside the score and never folded into it, so a relationship nobody
has published yet is not penalised for being new. It takes a local export
rather than a live query, which keeps the analysis offline, deterministic and
independent of a database version the result could not record.

```{r annotate, eval = FALSE}
known <- data.frame(source = "P1", target = "M1",
                    support = 0.9, database = "local STRING export")

annotate_evidence(results, known)
```

## Reading the results

`report()` turns a validation into something readable. The console format
leads with the verdict and flags only what needs attention:

```{r report, eval = FALSE}
report(validation, format = "console")
```

The default format writes a self-contained HTML document — styles, scripts
and plots are all inlined, so it opens with no network access and no CDN
dependency:

```{r report-html, eval = FALSE}
report(validation, file = "validation_report.html")
```

## Development

```{r dev, eval = FALSE}
devtools::document()
devtools::test()
devtools::check()
```

`tests/manual/` holds two standalone walkthroughs that drive the package the
way a user would and print everything they would see at the console.
`walkthrough_simple.R` runs on simulated data; `walkthrough_complex.R` runs
on the NHANES 1999-2006 exposome release, which is not redistributed here —
`tests/manual/test_data/SOURCE.md` says where to get it. Neither is part of
the built package.
