---
title: "From blocks of data to a claim you could be wrong about"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From blocks of data to a claim you could be wrong about}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CausalMultiOmics)
```

Most analysis packages fit a model and hand it back. This one does not,
because "the best model" is a poor answer to a biological question: the model
that predicts best is usually the one leaning hardest on whatever was
measured most accurately, which is a fact about the laboratory rather than
about the disease.

What comes out instead is a scored graph in which every arrow carries its own
evidence — which methods saw it, how large it was, how precisely it was
measured — and, the part that matters, what would have to be true for it to
mean what it appears to mean.

## Data with a known answer

Real data cannot show that an analysis is right. It has no answer to check
against. So this vignette plants one.

```{r simulate}
structure <- data.frame(
  from   = c("age", "age", "diet", "inflammation"),
  to     = c("diet", "disease", "inflammation", "disease"),
  effect = c(0.4, 0.5, 0.8, 0.7)
)

sim <- simulate_data(
  n = 400,
  blocks = list(
    clinical = c("diet", "inflammation"),
    proteins = 8
  ),
  dag = structure,
  outcome = "disease",
  missing = c(proteins = 0.1),
  seed = 1
)

sim
```

Read the structure as biology. Age influences both what people eat and
whether they die, so it confounds. Diet drives inflammation, which in turn
drives the disease — so inflammation is a **mediator**, and the eight
proteins are noise with nothing to do with any of it.

The truth travels with the data:

```{r truth}
sim$misc$simulation$planted_features
sim$misc$simulation$noise_features
```

## Audit before you touch anything

`check_data()` changes nothing. It profiles each block, benchmarks candidate
transformations, and derives a preprocessing recipe with its reasoning
attached.

```{r check}
quality <- check_data(sim)

quality$summary$n_blocks
quality$summary$total_missing_percent
```

Two overlap figures are reported, and the second is the one an analysis runs
on:

```{r overlap}
quality$summary$shared_by_all

quality$summary$cumulative_overlap
```

Every pair of twenty blocks can share the whole cohort while the twenty
together share none of it, because each can be missing a different part. A
pairwise matrix cannot express that; the running total can.

## Execute the plan, and nothing else

`preprocess()` derives nothing of its own. Every method, parameter and
threshold comes from the recipe, and every stage stores its fitted model so
the pipeline can be replayed on a second cohort.

```{r preprocess}
clean <- preprocess(sim, quality, plots = FALSE, quiet = TRUE)

clean$quality$table[, c("block", "missing_before", "missing_after")]
```

A block going from 10% missing to 0% did not improve. It was filled in, and
that matters later.

## Build evidence

```{r analyze}
results <- analyze(
  clean,
  outcome = "disease",
  covariates = "age",
  methods = c("association", "conditional"),
  effort = "standard",
  plots = FALSE,
  quiet = TRUE
)

results
```

The planted driver should be at the top, and the eight noise proteins should
not be:

```{r drivers}
head(results$interpretation$drivers[, c("source", "evidence_score",
                                        "identification")], 4)
```

Note the `identification` column. Every relationship records the strategy
that would license reading it causally, and `adjustment` means measured
confounders were handled and unmeasured ones cannot be. It is a hypothesis,
not a conclusion, however high it scores.

## What an adjustment is actually worth

Adjusting for a variable helps only when the variable is a confounder.
Adjusting for a mediator deletes part of the effect being measured; adjusting
for a collider manufactures an association out of nothing. Both look like
diligence and both make the estimate worse than doing nothing.

Nothing in the data tells these apart — the correlation matrix is identical
in all three cases. Only a structure does, and a structure is a claim about
the world that has to come from you.

```{r dag, eval = requireNamespace("dagitty", quietly = TRUE)}
check_dag(structure, "diet", "disease", adjusted = "age")
```

And the same question with the mediator wrongly included:

```{r dag-bad, eval = requireNamespace("dagitty", quietly = TRUE)}
check_dag(structure, "diet", "disease", adjusted = c("age", "inflammation"))
```

Passing `dag = structure` to `analyze()` audits every relationship this way.
It changes no estimate — a structure is an interpretive claim, not a
statistical one — but it changes what may be concluded, and an adjustment
that includes a mediator loses its label rather than keeping it with a
footnote.

## What was measured and what was invented

Imputation fills gaps with plausible numbers, which is usually right. But
once a gap is filled nothing downstream can tell a measurement from an
estimate, and the confidence interval comes out exactly as narrow as if every
value had been real.

```{r quality}
affected <- Filter(function(e) length(e$quality_flags) > 0, results$evidence)

length(affected)

if (length(affected) > 0) affected[[1]]$quality_flags
```

`data_quality` is the share of the data behind a relationship that was
measured rather than reconstructed. It scales the evidence score rather than
entering the confidence score, because those are different problems: a wide
interval is fixed by collecting more people and a column that was half
invented is fixed by nothing.

## Does one number describe everybody?

An average says nothing about whether the people it averages resemble each
other. A coefficient of 0.4 is equally compatible with 0.4 in everyone, and
with 2.0 in a tenth of them and nothing in the rest.

```{r spread}
checked <- Filter(function(e) is.finite(e$share_driving_effect),
                  results$evidence)

for (e in head(checked, 3)) {
  cat(sprintf("%-16s halved by removing %d people (%.0f%%)\n",
              e$source, e$samples_driving_effect,
              100 * e$share_driving_effect))
}
```

For a relationship that holds across a group the answer is most of them,
because removing a handful barely shifts an average. A small answer means the
finding is carried by a minority nobody has identified.

## State something that could be wrong

```{r hypothesis}
h <- hypothesis(results)

h$claim
h$grade
```

The grade comes from identification alone. Precision, method agreement and
resampling stability describe how *well* an association was estimated and say
nothing about what it is evidence *of*; letting them raise the grade would
turn a well-measured correlation into a cause by arithmetic.

The sentence follows the grade rather than the score. "Higher X goes with
higher Y" describes what was seen; "raising X would raise Y" describes what
would happen, and only identification licenses the second.

The part a result never has is what would settle the argument:

```{r settles}
h$settles
```

Each item comes from a specific weakness of this specific relationship rather
than from a stock paragraph of caution. A reader can act on "measure a
confounder associated with both at 3.2 or more". Nobody can act on "residual
confounding cannot be excluded".

## Take it somewhere new

A claim carries its own protocol — which outcome, which covariates, which
model — so what runs on a second cohort is what ran on the first.

```{r replicate}
validation <- simulate_data(
  n = 350,
  blocks = list(clinical = c("diet", "inflammation"), proteins = 8),
  dag = structure,
  outcome = "disease",
  seed = 2
)

tested <- test_hypothesis(h, apply_preprocessing(validation, clean,
                                                 quiet = TRUE))

tested$replication$verdict
tested$replication$ratio
```

Replication is judged on direction and size, not on a p-value. The usual way
one is oversold is a direction that holds with an effect a fifth as large,
which a significance test calls a success; that case is reported as
`"same direction, much smaller"`.

## Where to go next

- `report(results, file = "report.html")` writes a self-contained document
  intended for a reader who is not a statistician, with every finding
  carrying what would settle it.
- `explain(results, "diet")` gathers everything known about one variable,
  including what each method reported on its own before anything was merged.
- `counterfactual(results, modifiable = "clinical")` states the findings in
  the units the variables were measured in, and includes nothing you have not
  declared changeable.
- `results$consensus` and `results$modules` read the same evidence across
  resamples and at the resolution between a variable and a block.

Nothing here is offered as certainty. The package exists to make the gap
between what was measured and what may be claimed visible rather than
convenient.
