---
title: "From blocks of data to a claim you could be wrong about"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From blocks of data to a claim you could be wrong about}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(CausalMultiOmics)
```

Most analysis packages fit a model and hand it back. This one does not,
because "the best model" is a poor answer to a biological question: the model
that predicts best is usually the one leaning hardest on whatever was
measured most accurately, which is a fact about the laboratory rather than
about the disease.

What comes out instead is a scored graph in which every arrow carries its own
evidence — which methods saw it, how large it was, how precisely it was
measured — and, the part that matters, what would have to be true for it to
mean what it appears to mean.

## Data with a known answer

Real data cannot show that an analysis is right. It has no answer to check
against. So this vignette plants one.

```{r simulate}
structure <- data.frame(
  from   = c("age", "age", "diet", "inflammation"),
  to     = c("diet", "disease", "inflammation", "disease"),
  effect = c(0.4, 0.5, 0.8, 0.7)
)

sim <- simulate_data(
  n = 400,
  blocks = list(
    clinical = c("diet", "inflammation"),
    proteins = 8
  ),
  dag = structure,
  outcome = "disease",
  missing = c(proteins = 0.1),
  seed = 1
)

sim
```

Read the structure as biology. Age influences both what people eat and
whether they die, so it confounds. Diet drives inflammation, which in turn
drives the disease — so inflammation is a **mediator**, and the eight
proteins are noise with nothing to do with any of it.

The truth travels with the data:

```{r truth}
sim$misc$simulation$planted_features
sim$misc$simulation$noise_features
```

## Audit before you touch anything

`check_data()` changes nothing. It profiles each block, benchmarks candidate
transformations, and derives a preprocessing recipe with its reasoning
attached.

```{r check}
quality <- check_data(sim)

quality$summary$n_blocks
quality$summary$total_missing_percent
```

Two overlap figures are reported, and the second is the one an analysis runs
on:

```{r overlap}
quality$summary$shared_by_all

quality$summary$cumulative_overlap
```

Every pair of twenty blocks can share the whole cohort while the twenty
together share none of it, because each can be missing a different part. A
pairwise matrix cannot express that; the running total can.

## Execute the plan, and nothing else

`preprocess()` derives nothing of its own. Every method, parameter and
threshold comes from the recipe, and every stage stores its fitted model so
the pipeline can be replayed on a second cohort.

```{r preprocess}
clean <- preprocess(sim, quality, plots = FALSE, quiet = TRUE)

clean$quality$table[, c("block", "missing_before", "missing_after")]
```

A block going from 10% missing to 0% did not improve. It was filled in, and
that matters later.

## Build evidence

```{r analyze}
results <- analyze(
  clean,
  outcome = "disease",
  covariates = "age",
  effort = "standard",
  control = analysis_control(methods = c("association", "conditional")),
  plots = FALSE,
  quiet = TRUE
)

results
```

The planted driver should be at the top, and the eight noise proteins should
not be:

```{r drivers}
head(results$interpretation$drivers[, c("source", "evidence_score",
                                        "identification")], 4)
```

Note the `identification` column. Every relationship records the strategy
that would license reading it causally, and `adjustment` means measured
confounders were handled and unmeasured ones cannot be. It is a hypothesis,
not a conclusion, however high it scores.

## What an adjustment is actually worth

Adjusting for a variable helps only when the variable is a confounder.
Adjusting for a mediator deletes part of the effect being measured; adjusting
for a collider manufactures an association out of nothing. Both look like
diligence and both make the estimate worse than doing nothing.

Nothing in the data tells these apart — the correlation matrix is identical
in all three cases. Only a structure does, and a structure is a claim about
the world that has to come from you.

```{r dag, eval = requireNamespace("dagitty", quietly = TRUE)}
check_dag(structure, "diet", "disease", adjusted = "age")
```

And the same question with the mediator wrongly included:

```{r dag-bad, eval = requireNamespace("dagitty", quietly = TRUE)}
check_dag(structure, "diet", "disease", adjusted = c("age", "inflammation"))
```

Passing `dag = structure` to `analyze()` audits every relationship this way.
It changes no estimate — a structure is an interpretive claim, not a
statistical one — but it changes what may be concluded, and an adjustment
that includes a mediator loses its label rather than keeping it with a
footnote.

## What was measured and what was invented

Imputation fills gaps with plausible numbers, which is usually right. But
once a gap is filled nothing downstream can tell a measurement from an
estimate, and the confidence interval comes out exactly as narrow as if every
value had been real.

```{r quality}
affected <- Filter(function(e) length(e$quality_flags) > 0, results$evidence)

length(affected)

if (length(affected) > 0) affected[[1]]$quality_flags
```

`data_quality` is the share of the data behind a relationship that was
measured rather than reconstructed. It scales the evidence score rather than
entering the confidence score, because those are different problems: a wide
interval is fixed by collecting more people and a column that was half
invented is fixed by nothing.

## Does one number describe everybody?

An average says nothing about whether the people it averages resemble each
other. A coefficient of 0.4 is equally compatible with 0.4 in everyone, and
with 2.0 in a tenth of them and nothing in the rest.

```{r spread}
checked <- Filter(function(e) is.finite(e$share_driving_effect),
                  results$evidence)

for (e in head(checked, 3)) {
  cat(sprintf("%-16s halved by removing %d people (%.0f%%)\n",
              e$source, e$samples_driving_effect,
              100 * e$share_driving_effect))
}
```

For a relationship that holds across a group the answer is most of them,
because removing a handful barely shifts an average. A small answer means the
finding is carried by a minority nobody has identified.

## State something that could be wrong

```{r hypothesis}
h <- hypothesis(results)

h$claim
h$grade
```

The grade comes from identification alone. Precision, method agreement and
resampling stability describe how *well* an association was estimated and say
nothing about what it is evidence *of*; letting them raise the grade would
turn a well-measured correlation into a cause by arithmetic.

The sentence follows the grade rather than the score. "Higher X goes with
higher Y" describes what was seen; "raising X would raise Y" describes what
would happen, and only identification licenses the second.

The part a result never has is what would settle the argument:

```{r settles}
h$settles
```

Each item comes from a specific weakness of this specific relationship rather
than from a stock paragraph of caution. A reader can act on "measure a
confounder associated with both at 3.2 or more". Nobody can act on "residual
confounding cannot be excluded".

## Take it somewhere new

A claim carries its own protocol — which outcome, which covariates, which
model — so what runs on a second cohort is what ran on the first.

```{r replicate}
validation <- simulate_data(
  n = 350,
  blocks = list(clinical = c("diet", "inflammation"), proteins = 8),
  dag = structure,
  outcome = "disease",
  seed = 2
)

tested <- test_hypothesis(h, apply_preprocessing(validation, clean,
                                                 quiet = TRUE))

tested$replication$verdict
tested$replication$ratio
```

Replication is judged on direction and size, not on a p-value. The usual way
one is oversold is a direction that holds with an effect a fifth as large,
which a significance test calls a success; that case is reported as
`"same direction, much smaller"`.

## Two kinds of argument

`analyze()` takes its many options in two groups, because they are two kinds
of decision.

`analysis_assumptions()` holds what you are claiming: the causal diagram,
which variables are modifiable, the reliability of an assay, which events
compete with the outcome. None of it can be derived from the data, all of it
changes what may be concluded, and getting one wrong makes the answer wrong.

`analysis_control()` holds how much work to do: which generators, how many
resamples, how many features. Getting one wrong makes the answer slower or
noisier, never wrong.

```{r groups, eval = FALSE}
analyze(
  clean,
  outcome = "disease",
  covariates = "age",
  assume = analysis_assumptions(
    dag = structure,
    reliability = c(inflammation = 0.85),
    modifiable = "clinical"
  ),
  control = analysis_control(
    methods = c("association", "survival"),
    max_features = 200
  )
)
```

Two of those deserve a word.

`reliability` corrects for the attenuation that measurement error causes: a
noisy exposure pulls its own coefficient toward zero by exactly its
reliability. The correction is reported beside the measured estimate and
never in place of it, and it widens the interval rather than narrowing it —
an attenuation correction that left the interval alone would turn a noisy
measurement into a stronger claim.

`positivity`, on by default, checks the identification condition nobody
mentions. Reading an adjusted estimate causally needs two things: no
unmeasured confounding, and variation in the exposure at every combination of
the adjustment set. Where the second fails the model does not complain, it
extrapolates.

```{r positivity}
checked <- Filter(function(e) !is.na(e$positivity), results$evidence)

for (e in head(checked, 3)) {
  cat(sprintf("%-14s positivity %-5s (%.0f%% of it free to vary)\n",
              e$source, e$positivity, 100 * e$residual_variation))
}
```

Only relationships with the outcome are checked, since only those have an
adjustment set to be trapped by.

## Several outcomes at once

Studies rarely have one endpoint. Naming several runs each analysis on its
own terms and then corrects multiplicity across all of them together — which
running `analyze()` twice by hand does not, because a relationship that was
one of fifty tests is not one of a hundred, and correcting within each
outcome separately pretends the other analysis was never run.

```{r multi, eval = FALSE}
analyze(clean, outcome = c("disease", "biomarker"), effort = "fast")
```

## Where to go next

- `report(results, file = "report.html")` writes a self-contained document
  intended for a reader who is not a statistician, with every finding
  carrying what would settle it.
- `explain(results, "diet")` gathers everything known about one variable,
  including what each method reported on its own before anything was merged.
- `counterfactual(results, modifiable = "clinical")` states the findings in
  the units the variables were measured in, and includes nothing you have not
  declared changeable.
- `results$consensus` and `results$modules` read the same evidence across
  resamples and at the resolution between a variable and a block.

Nothing here is offered as certainty. The package exists to make the gap
between what was measured and what may be claimed visible rather than
convenient.
